Limits to Ecological Forecasting: Estimating Uncertainty for Critical Transitions with Deep Learning

### 2022-08-10 manuscript/manuscript.Rmd

In this paper, we carry out an initial exploration on how deep learning methods can perform on predicting critical transition events.
We compare the ability of several cutting edge machine learning approaches against statistical and process-based models, and show that deep learning methods are generally able to strike a middle ground between what we consider as acceptable and ideal case forecasting methods, ARIMA and MCMC-based parameter estimation respectively.
Although most ML-based forecasting applications focus on point predictions, we have emphasized examples that can provide informative estimates of uncertainty.
When the ML models are able to observe transition phenomena, as in the stochastic case, they performed comparably to MCMC-based forecasting with respect to CRPS and log probability score but under-performed MCMC when there were no transition events in the training sets as in the Hopf and saddle-node examples.
An ensemble forecast combining the predictions of all four ML methods generally scores as well or better than any one of the ML methods alone.
Yet, examining summary statistics, CRPS and log probability scores obscures finer detailed components of the forecasts.
For instance, forecast skill varies with the length of the forecast horizon in a non-monotonic fashion. 
This is the result of multiple factors: for some dynamics, such as those involving tipping points, the long term behavior can be easier to predict than transient transitions. 
Both predicted uncertainty and forecast skill can be better on longer horizons than on shorter ones, as in the MCMC predictions of tipping point dynamics.
It is also important to remember that probabilistic forecast skill scores do not only measure how close observations are to expectation, but also reflect the predicted uncertainty: therefore, over-confidence about predictive accuracy can result in worse scores than the scores of forecasts that are less accurate on average but correctly reflect a greater degree of uncertainty.
The ability to better reflect uncertainty rather than better average predictions explains much of the performance of the ensemble model.

@@

When examining the performance of the ML models, it is clear that there is no single method that excels in all scenarios.
Neither is there one class of ML methods that outperforms the others -- a fact we found surprising given the reported dominance of encoder-decoders in the sequence-to-sequence deep learning sub-field [@aitken_understanding_2021].
These observations underscore the point that ML is a very empirically-driven field in which there are few guarantees on performance.
Furthermore, due to the black-box-ness of deep learning and other reasons like instability to initialization seed, it is often impossible to provide an explanation for why certain methods over-perform or fail to meet expectations.

@@

It may be natural to ask whether this performance would be remedied if the ML models were trained on data which includes prior examples of super-critical Hopf or saddle node bifurcations. 
This question is not as easy to answer as it may seem, because of the difficulty in defining the corresponding forecasting scenario.
The scenarios we have considered are true, pure forecasts: the training data comes from a single realization of a specific generative process, and the task is to predict the future states of that system before they occur.
Would it be possible to train a predictive algorithm on 'analogous' examples of critical transitions?
For instance, could data from other lakes which may have experienced a critical transition such as eutrophication event in the past be used to train machine learning models to predict such events in some focal lake in the future [@Scheffer2001]?
Perhaps, but it depends what we mean by an 'analogous' system.
Even if the underlying mechanism was accurately captured by the same model, say, the saddle-node model of @May1977 we consider here, it is likely that most of the individual model parameters would be quite different, even after accounting for re-scaling or non-dimensionalization of the model [@Hastings1996].
Rarely do ecologists have access to completely controlled replicates for fitting or training models.
The ability for ML models to successfully generalize from training in such cases remains an open problem and a promising subject of further investigation.

There are a number of questions that we have left unanswered that we hope will be addressed in future work.
In this paper, we have explored a small number of machine learning and statistical models that can be used for forecasting, so comprehensive conclusions can not be drawn on whether statistical or machine learning-based approaches are better suited for tipping point forecasting problems.
Neither can we claim that the ML methods employed will translate well to all sudden transition event forecasting problems in reality, since working with real data will introduce additional difficulties like how to deal with missing data, sparse data and observation errors.

@@

Ecological forecasting is invariably difficult, even in the idealized cases of ample measurement data and clearly identified structural models.
This paper is not intended to give a complete answer to whether deep learning is the best suited method for tipping point forecasting problems as this will take numerous studies to resolve; instead, this study aims to be an early exploration on whether deep learning methods should be considered as viable tools for this extremely challenging class of prediction problems. <!-- I think here would perhaps be a good place to flesh out "promising" tag, stuck on how to do so-->
Since we have shown that ML models can make good forecasts on our selected tipping point scenarios, particularly the stochastic transition case, we have established that these deep learning models are promising methods for critical transition forecasting problems.

>>>>

In this paper, we carry out an initial exploration on how deep learning methods can perform on predicting critical transition events.
We compare the ability of several cutting edge machine learning approaches against statistical and process-based models, and show that deep learning methods are generally able to strike a middle ground between what we consider as acceptable and ideal case forecasting methods, ARIMA and MCMC-based parameter estimation respectively.
Although most ML-based forecasting applications focus on point predictions, we have emphasized examples that can provide estimates of uncertainty.
When the ML models are able to observe transition phenomena, as in the stochastic case, they performed comparably to MCMC-based forecasting with respect to CRPS and log probability score but under-performed MCMC when there were no transition events in the training sets as in the Hopf and saddle-node examples.
An ensemble forecast combining the predictions of all four ML methods generally scores as well or better than any one of the ML methods alone.
Yet, examining summary statistics, CRPS and log probability scores obscures finer detailed components of the forecasts.
For instance, forecast skill varies with the length of the forecast horizon in a non-monotonic fashion. 
This is the result of multiple factors: for some dynamics, such as those involving tipping points, the long term behavior can be easier to predict than transient transitions. 
Both predicted uncertainty and forecast skill can be better on longer horizons than on shorter ones, as in the MCMC predictions of tipping point dynamics.
It is also important to remember that probabilistic forecast skill scores do not only measure how close observations are to expectation, but also reflect the predicted uncertainty: therefore, over-confidence about predictive accuracy can result in worse scores than scores from forecasts that are less accurate on average but correctly reflect a greater degree of uncertainty.
The ability to better reflect uncertainty rather than better average predictions explains much of the performance of the ensemble model.

@@

When examining the performance of the ML models, it is clear that there is no single method that excels in all scenarios.
Neither is there one class of ML methods that outperforms the others -- a fact we found surprising given the reported dominance of encoder-decoders in the field of sequence-to-sequence deep learning [@aitken_understanding_2021].
These observations underscore the point that ML is a very empirically-driven field in which there are few guarantees on performance.
Furthermore, due to the black-box-ness of deep learning and other reasons like instability to initialization seed, it is often impossible to provide an explanation for why certain methods over-perform or fail to meet expectations.

@@

It may be natural to ask whether this performance would be remedied if the ML models were trained on data which includes prior examples of super-critical Hopf or saddle node bifurcations. 
This question is not as easy to answer as it may seem, because of the difficulty in defining the corresponding forecasting scenario.
The scenarios we have considered are true, pure forecasts: the training data comes from a single realization of a specific generative process, and the task is to predict the future states of that system before they occur.
Would it be possible to train a predictive algorithm on 'analogous' examples of critical transitions?
For instance, could data from other lakes, which may have experienced a critical transition such as an eutrophication event in the past, be used to train machine learning models to predict such events in some focal lake in the future [@Scheffer2001]?
Perhaps, but it depends on what we mean by an 'analogous' system.
Even if the underlying mechanism was accurately captured by the same model, say, the saddle-node model of @May1977 we consider here, it is likely that most of the individual model parameters would be quite different, even after accounting for re-scaling or non-dimensionalization of the model [@Hastings1996].
Rarely do ecologists have access to completely controlled replicates for fitting or training models.
The ability for ML models to successfully generalize from training in such cases remains an open problem and a promising subject of further investigation.

There are a number of questions that we have left unanswered that we hope will be addressed in future work.
In this paper, we have explored a small number of machine learning and statistical models that can be used for forecasting, so comprehensive conclusions can not be drawn on whether statistical or machine learning-based approaches are better suited for critical transition forecasting problems.
Neither can we claim that the ML methods employed will translate well to all sudden transition event forecasting problems in reality, since working with real data will introduce additional difficulties like how to deal with missing data, sparse data and observation errors.

@@

Ecological forecasting is invariably difficult, even in the idealized cases of ample measurement data and clearly identified structural models.
This paper is not intended to give a complete answer to whether deep learning is the best suited method for tipping point forecasting problems as this will take numerous studies to resolve; instead, this paper aims to be an early exploration on whether deep learning methods should be considered as viable tools for this extremely challenging class of prediction problems. <!-- I think here would perhaps be a good place to flesh out "promising" tag, stuck on how to do so-->
Since we have shown that ML models can make relatively good forecasts on our selected tipping point scenarios, particularly the stochastic transition case, we have established that these deep learning models are worthy of further investigation regarding their application to critical transition forecasting problems.
